---
title: "Adaptive Mobile: coverType Data Set - Exploratory Analysis "
author: "Brian Carter"
date: "24 March, 2015"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_width: 6
    fig_height: 4
    fig_caption: false
---

```{r setup, eval=TRUE,echo=FALSE}

setwd("C:\\Users\\IBM_ADMIN\\Desktop\\github\\adaptm\\DataExplore_with_R")

#Set some global options for r chunks. Want R chunk to evaluate, but not included for client facing document. 
#knitr::opts_chunk$set(cache=TRUE,cache.lazy = TRUE,
#												fig.height=4,fig.width=6,fig.align='center',fig.path='figure/',
#											warning=FALSE,message=FALSE,error=FALSE,echo=FALSE
#										)  ##background doesn't work
#

knitr::opts_chunk$set(cache=FALSE,cache.lazy = FALSE,
											fig.path='figure/',
											warning=FALSE,message=FALSE,error=FALSE,echo=FALSE
										)  ##background doesn't work


#Caching gives some funny results but it can result in faster loading times, when writing up the document. This is to do with the := assignment in data.table. Knitr doesn't recognise, therefore always use <-

```

```{r,cache=FALSE}

##Libraries required
#Data work 
require(data.table) #Working with large files
require(xlsx)       #Loading and saving .xlsx files 
require(plyr)   #Always load in this order plyr, dpply, lubridate - dpplyr overrides some of methods in plyr. 
require(dplyr) #Use require as it will give an error message if the package doesn't exist
require(lubridate) #used for working with data information. 
require(reshape2)  #used for melting 
require(ggplot2)
#install.packages("ggthemes")
require(ggthemes)
require(scales)
require(GGally)
#require(cowplot)
require(gridExtra)

#Formating and printing 
#install.packages("devtools")
#devtools::install_github("adletaw/captioner")   #Nice library for numbering and captioning tables in conjunction with knitr and pandoc
require(pander)   	#for creating nice output tables.
require(captioner)

#Set up the figure and table numbering
fig_nums<-captioner(prefix = "Figure")
tab_nums<-captioner(prefix = "Table")

#Using pryr abbreviate how to call fig_nums function 
require(pryr)
citefig<-pryr::partial(fig_nums,display="cite")
citetab<-pryr::partial(tab_nums,display="cite")

#Turn off caption.prefix as allow captioner to handle this. 
panderOptions('table.caption.prefix', '')
panderOptions('big.mark', ",")
panderOptions('keep.trailing.zeros', TRUE)
panderOptions('keep.line.breaks',TRUE)
panderOptions('table.emphasize.rownames',TRUE)


#Create theme for ggplots
theme_adMobile <- function(){
     theme(
      axis.text.x=element_text(angle=30,hjust=1,size=8),
      axis.text.y=element_text(size=8),
      axis.title.x=element_text(size=10),
      axis.title.y=element_text(size=10),
      panel.background = element_blank()
      )
}
```

# Excutive Summary


NOTCOMPLETE 

This document present an exploration of the data contained [Forest CoverType dataset, College of Natural Resources, Colorado State University](https://archive.ics.uci.edu/ml/datasets/Covertype). An introduction to the information contained in the dataset is presented. The original dataset is described and additional metadata is added for the purposes of more intuitive understanding of the data. 

The features of the dataset are analyised. This is performed in the R language. Through examine the data a number of research questions are presented to be answered for the purposes of. EXPAND HERE. 

The research questions are answered in Python. 


# Introduction

```{r readdata,cache=FALSE,results='hide'}
##File for the data file. 
theFile="C:\\Users\\IBM_ADMIN\\Desktop\\OneDrive\\PublicData\\AM\\v1\\covtype.data"

#Read in the file, try to use the syntax as applied here to keep track of what packages do what. 
coverData<-data.table::fread(theFile,header=FALSE,na.strings=c(""))
```

The dataset in its original form contains `r nrow(coverData)` rows across `r ncol(coverData)` columns. The dataset doesn't contain column headers. These are contained in the meta data file **covtype.info** file. These are added to the data. (An excel file with the meta data was created)

```{r colheads,cache=FALSE,results='hide',include=FALSE}
#Add the column headers from created Excel file
require(xlsx)
theFile="C:\\Users\\IBM_ADMIN\\Desktop\\github\\adaptm\\metadata.xlsx"
colNames<-data.table(read.xlsx(theFile, sheetName = "colNames"))
coverData<-setnames(coverData, as.character(colNames[,dbName]) )
```

XXXXDelete Kaggle Names in metdata.xlsxXXXX
XXXXSave with column names?XXXX

```{r checkmutual,cache=FALSE,results='hide',include=FALSE}
#Check for mutually exclusive columns 
coverData<-coverData[, testWA := rowSums(.SD), .SDcols = 11:14]
coverData<-coverData[,testST := rowSums(.SD), .SDcols = 15:54]

coverData[testWA>1,]  #zero returned
coverData[testST>1,]  #zero returned
sum(coverData$testWA)==nrow(coverData)  #True
sum(coverData$testST)==nrow(coverData)  #True

#Drop the colums
coverData<-coverData[,testWA := NULL]
coverData<-coverData[,testST := NULL]
```


Columns 1-10 are quantiative data. Columns 11-14 are a binary resentation of the four possible types of **Wilderness_Area** with a 1 representing the presence of that wilderness type and 0 otherwise. These columns are mutually exclusive, in that, for each row only only of the four columns can contain a 1. For the purpose of exploratory analysis the four columns are reduced to their **original** representation of 1 column. 

Similarly columns 15-54 are a binary representation of the 40 possible **Soil_Types**. For the purposes of exploratory analysis, these 40 columns are reduced to their original state with one column REWRITE. 

```{r reducebinary,cache=FALSE,results='hide',include=FALSE}
#Reduce the two binary variable to original state with variables and encode the coverType variable with original labels. 

#Recode coverType with character labels for exploratory analysis. 
metaData <- data.table(read.xlsx(theFile,sheetName = "targetName"))

setkeyv(coverData,"coverType")
setkeyv(metaData,"value")
coverData <- coverData[metaData,coverType2:=i.newLabel]
coverData <- coverData[,coverType:=NULL]
setnames(coverData,"coverType2","coverType")

#Read in the labels for Wilderness Area and use for loop to update the data.table
metaData <-data.table(read.xlsx(theFile, sheetName = "wildernessArea"))

for(i in 1:nrow(metaData)) {
   refColumn<-as.character(metaData[i,dbName])
   refValue<-as.character(metaData[i,newLabel])
   coverData<-coverData[get(refColumn)==1,wildernessArea:=refValue]
}

#Read in the labels for Wilderness Area and use for loop to update the data.table
metaData <-data.table(read.xlsx(theFile, sheetName = "soilType"))

for(i in 1:nrow(metaData)) {
   refColumn<-as.character(metaData[i,dbName])
   refValue<-as.character(metaData[i,newLabel])
   coverData<-coverData[get(refColumn)==1,soilType:=refValue]
}

#remove the binary columns
coverData <- coverData[ , colnames(coverData[,11:54,with=FALSE]):=NULL]

```

After reducing the binary columns there are ` r ncol(coverData)` columns in the dataset. In the **covtype.info** file there is extra meta data associated with the **soilType** column that is not in the original dataset. Each **soilType** has an unique associated 4-digit *ELU.Code*. The first digit of the *ELU.Code* represents its' *Climatic Zone* and the second digit its' *Geological Zone*. This extra information is introduced for exploratory analysis. 

```{r targetlabel,cache=FALSE,results='hide',include=FALSE}
#Add some additional information contained about soil type. 

setkeyv(coverData,"soilType")
setkeyv(metaData,"newLabel")

coverData<-coverData[metaData,climaticZone :=i.climatic.zone]
coverData<-coverData[metaData,geologicZone :=i.geologic.zones]

#reorder the columns
colOrder<-c("Elevation","Aspect","Slope","HD.Hydro","VD.Hydro","HD.Road","HD.Fire","HS.9am","HS.noon","HS.3pm","wildernessArea","soilType","climaticZone","geologicZone","coverType")  

setcolorder(coverData, colOrder)

#Remove some variables
rm(i,colOrder,refColumn,refValue,metaData,theFile,colNames)
```

The final column represents the **coverType** with seven possible values. **coverType** has been the main predictive target in much analysis concerning this dataset. The lables for this column are used during exploratory analysis. 

## Data Summary

Having reduced the binary columns representations of  **Wilderness_Area, Soil_Type** to their original form and added two additional columns there are `r ncol(coverData)` columns in the dataset. 


```{r,results='hide'}

#summaryTable to gather statistics. 
summaryTable<-data.table(name=names(coverData))

#May add description here? 
summaryTable<-summaryTable[,dataType:= sapply(coverData,class)]  
summaryTable<-summaryTable[,missing := t(coverData[,lapply(.SD,function(x) length(which(is.na(x))))]),]
summaryTable<-summaryTable[,unique :=  t(coverData[,lapply(.SD,function(x) length(unique(x)))]),]

integerCols<-summaryTable[dataType=="integer",name]
categoricalCols<-summaryTable[dataType!="integer",name]

tempCoverData<-coverData[,integerCols,with=FALSE]

intSummary<-data.table(name=names(tempCoverData))

measuredIn<-c("meters","azimuth","degrees","meters","meters","meters","meters","0-255","0-255","0-255")
intSummary<-intSummary[,quantity := measuredIn]

intSummary<-intSummary[,min :=  t(tempCoverData[,lapply(.SD,function(x) min(x))]),]
intSummary<-intSummary[,max :=  t(tempCoverData[,lapply(.SD,function(x) max(x))]),]
intSummary<-intSummary[,mean :=  t(tempCoverData[,lapply(.SD,function(x) round(mean(x),2))]),]
intSummary<-intSummary[,std :=  t(tempCoverData[,lapply(.SD,function(x) round(sd(x),2))]),]

summaryTable<-merge(summaryTable,intSummary,by="name",all=TRUE,sort=FALSE)
#summaryTable[is.na(summaryTable)] <- " "
```

```{r,results='hide'}
mytable<-data.frame(summaryTable)
tab_nums("sumTab","Summary of CoverTye Dataset Featuers")
```

A summary of the `r ncol(coverData)` is presented in `r citetab("sumTab")`. 

```{r,results='asis'}
pandoc.table(mytable,caption=tab_nums("sumTab"),justify=c("left"),missing="")
```

XXXXFix table if have time?XXXX

# Data Exploration

The **covtype.info** file includes refences to a number of papers where predictive models were built with **coverType** as the target variable. 



```{r,results='hide'}
#tempTable using dplyr
coverType.cnt = coverData %>% 
  group_by(coverType) %>% 
  summarise(count=n()) %>% 
  mutate(pct=count/sum(count)) 

plot<-ggplot(coverType.cnt, aes(x=coverType, y=count)) +
  geom_bar(stat="identity") +
  scale_y_continuous(labels=comma) +
  geom_text(data=coverType.cnt, aes(label=count,y=count+30000),size=3) +
  geom_text(data=coverType.cnt, aes(label=paste0(round(pct*100,1),"%"),y=count+10000),size=3) +
  theme_adMobile()

fig_nums("countCT","Distribution of coverType")
```

*coverType* is not evenly distributed. Two classes *(SpruceFir, LodgepolePine)* comprimise 85% of all rows. `r citefig("countCT")` displays a bar graphy of *coverType*.

```{r figure1,echo=FALSE}
plot+ggtitle(eval(fig_nums("countCT")))
```

## Numerical Data Exploration

Density Plots

```{r,results='hide'}
featuresToPlot<-c("Elevation","Aspect","Slope","HD.Hydro","VD.Hydro","HD.Road","HD.Fire","HS.9am","HS.noon","HS.3pm")
p=list()

for(i in 1:length(featuresToPlot)){
  p[[i]] <- ggplot(coverData, aes_string(x=featuresToPlot[i])) + 
              geom_density() + 
              theme_adMobile() + 
              theme(axis.title.y=element_blank())
  }
fig_nums("denseNumeric","Density Plot of Numeric Data")
```

```{r,figure2,echo=FALSE}
do.call(grid.arrange,c(p,top=eval(fig_nums("denseNumeric"))))
```

Boxplots


```{r,results='hide'}
featuresToPlot<-c("Elevation","Aspect","Slope","HD.Hydro","VD.Hydro","HD.Road","HD.Fire","HS.9am","HS.noon","HS.3pm")
p=list()
for(i in 1:length(featuresToPlot)){
  p[[i]] <- ggplot(coverData, aes_string(y=featuresToPlot[i],x="coverType")) + 
              geom_boxplot(outlier.shape = NA) + 
              theme_adMobile() +
              theme(axis.text.x=element_blank(),axis.title.x=element_blank())
    
  }
fig_nums("boxNumeric","Boxplot of Numeric Data - coverType")
```

```{r,figure3,echo=FALSE}
do.call(grid.arrange,c(p,top=eval(fig_nums("boxNumeric"))))
```

Correlation Matrix


```{r,results='hide'}
plot<-ggcorr(coverData[,1:10,with=FALSE], label = TRUE, label_size = 3, label_round = 2, label_alpha = TRUE, hjust = 00.75, size = 3,layout.exp = 1)
fig_nums("corrTen","Correlation Matrix of 10 Numerical Features")
```

```{r figure4,echo=FALSE}
plot+ggtitle(eval(fig_nums("corrTen")))
```


ScatterPlots

```{r,results='hide'}
corrFeature1<-c("HS.9am","HD.Fire","HD.Hydro","Slope","Aspect","Aspect")
corrFeature2<-c("HS.noon","HS.noon","VD.Hydro","HS.9am","HD.Fire","HS.noon")

scatterTemp<-sample_n(coverData,10000)

p=list()
for(i in 1:length(corrFeature1)){
  p[[i]] <- ggplot(scatterTemp, aes_string(x=corrFeature1[i],y=corrFeature2[i])) +
              geom_point(alpha=1/10) +
              theme_adMobile()  
  }
fig_nums("scatterPlot6","Scatterplot of 6 Correlated Features")
```


```{r,figure5,echo=FALSE}
do.call(grid.arrange,c(p,top=eval(fig_nums("scatterPlot6"))))
```







## Categorical Data Exploration

```{r,results="hide"}

#Create four seperate plots and arrange
categoricalCols<-c("wildernessArea","soilType","climaticZone","geologicZone")

#Create row counts using dplyr
categoryCount <- coverData %>%group_by(wildernessArea) %>% summarise(count=n())
p1 <- ggplot(categoryCount, aes(x=reorder(wildernessArea, -count), y=count)) + 
        geom_bar(stat="identity") +
        xlab("wildernesArea") +
        scale_y_continuous(labels=comma) +
         theme_adMobile() + 
         theme(axis.title.y=element_blank())

categoryCount <- coverData %>%group_by(soilType) %>% summarise(count=n())
p2 <- ggplot(categoryCount, aes(x=reorder(soilType, -count), y=count)) + 
        geom_bar(stat="identity") +
        xlab("soilType") +
        scale_y_continuous(labels=comma) +
        theme_adMobile() +
         theme(axis.text.x=element_text(size=6),axis.title.y=element_blank()) 

categoryCount <- coverData %>%group_by(climaticZone) %>% summarise(count=n())
p3 <- ggplot(categoryCount, aes(x=reorder(climaticZone, -count), y=count)) + 
        geom_bar(stat="identity") +
        xlab("climaticZone") +
        scale_y_continuous(labels=comma) +
        theme_adMobile() +
        theme(axis.title.y=element_blank())

#Create row counts using dplyr
categoryCount <- coverData %>%group_by(geologicZone) %>% summarise(count=n())
p4 <- ggplot(categoryCount, aes(x=reorder(geologicZone, -count), y=count)) + 
        geom_bar(stat="identity") +
        xlab("geologicZone") +
        scale_y_continuous(labels=comma) +
        theme_adMobile() + 
        theme(axis.title.y=element_blank())

fig_nums("category4","Row Counts of Categorioal Features")
```


```{r,figure6,echo=FALSE}
grid.arrange(p1,p2,p3,p4,ncol=2,top=eval(fig_nums("category4")))
```

```{r,results='hide',eval=FALSE}
#Calculate some feature importance. Note save to .csv to save time. rf

require('FSelector')

#Calculate chi-square feature importane, convert to sorted data.table
chiSquare <- chi.squared(coverType~. ,coverData) #discrete target
chiSquare$features1 <- row.names(chiSquare)
colnames(chiSquare)[1] <- "chi.square"
chiSquare$chi.square <- round(chiSquare$chi.square,3)
chiSquare <- chiSquare[with(chiSquare, order(-chi.square)), ]


require('caret')
#Use a sample of the data for the rf calculation as it takes long time. 
sampleIndex <- createDataPartition(coverData$coverType, p = .1,list = FALSE,times = 1)
rf<- random.forest.importance(coverType~. ,coverData[ sampleIndex,], importance.type = 1) #1=mean decrease in accuracy, discrete target

rf$features2 <- row.names(rf)
colnames(rf)[1] <- "random.forest"
rf$random.forest<-round(rf$random.forest/100,3)
rf <- rf[with(rf, order(-random.forest)), ] 

rfcopy<-rf
chicopy<-chiSquare

#from dplyr
twoFI<-bind_cols(chiSquare,rf)
colOrder<-c("features1","chi.square","features2","random.forest")  
setcolorder(twoFI, colOrder)
write.csv(twoFI, file = "featureImportance_coverData.csv",row.names=FALSE)



#Maybe save to .csv to save time in future

```

```{r,results='hide'}
mytable<-read.csv(file="featureImportance_coverData.csv",header=TRUE)
tab_nums("featureImportTab","Feature Importance (chi,randomForest) of CoverTye Dataset Featuers")
```

FEATURE IMPORTANCE EXPLAIN

```{r,results='asis'}
pandoc.table(mytable,caption=tab_nums("featureImportTab"),justify=c("left"),missing="")
```

```{r,results='hide'}
#theFile="/Users/briancarter1/Desktop/OneDrive/PublicData/AM/v2/covtype15.csv"
#write.csv(coverData,file=theFile,row.names=FALSE)
```


# Summary

# Research Questions of Interest


